---
title: "scipt_clearing"
author: "Cliff Siu"
date: "2022-08-24"
output: html_document
---



## Install and load the packages
```{r}
install.packages("tidyverse")
install.packages("dplyr")
install.packages("ggplot2")

library(stringr)
library(tidyverse)
library(dplyr)
library(ggplot2)
```



## Load the data
```{r}
total_trip_clean<-read.csv("total_trip_clean.csv")

glimpse(total_trip_clean)
```




## Create a new order for data
I want to have a different arrangements for the day_of_week and months. So that I can work easier later.

```{r}
total_trip_clean$day_of_week <- 
  ordered(total_trip_clean$day_of_week, levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday'))


total_trip_clean$month <-
  ordered(total_trip_clean$month, levels=c(8,9,10,11,12,1,2,3,4,5,6,7))
```

###### change the column name for better display and understanding.
```{r}
colnames(total_trip_clean)[which(names(total_trip_clean) == "member_casual")] <- "type_of_rider"
```




## Analyze 

### Summary of the data with tables.

Here, we are looking at the following things:
* The average amount of riders for casual and member by the day of week and month.
* The average riding duration for casual riders and members riders by the day of week and month.
* The difference between two types of bikes, electric bikes and non-electric bikes by the day of week and month.
* The population of each station, and find out which one is the most popular.

#### The summary of riding duration in minutes between casual riders and member riders. 
```{r}
num_duration_v1 <- total_trip_clean %>%
  drop_na(type_of_rider) %>% 
  group_by(type_of_rider) %>%  
  summarise(number_of_rides = n(),
            lower = min(duration_mins),
            upper = max(duration_mins),
            average = mean(duration_mins),
            Median = median(duration_mins),
            SD = sd(duration_mins),
            range = max(duration_mins) - min(duration_mins)) %>% 
  view()
```
As we can see, the variance (variance  = SD^2) is so high. That the data points are very spread out from the mean.
And based on the range, we can tell there are quite a lot of outlines/ bad data. Further investigation is needed. 
For now, I only focus on the duration of a day. On the hand, I am going to remove anything under 1 minute, because it might cause by technical problems when the rider rented the bikes. Since it is not my interest in the case study and investigation is needed in the future. 


```{r}
num_duration_v2 <- total_trip_clean %>% 
  filter(duration_mins >1 & duration_mins < 1440) %>% 
  drop_na(type_of_rider) %>% 
  group_by(type_of_rider) %>%  
  summarise(number_of_rides = n(),
            lower = min(duration_mins),
            upper = max(duration_mins),
            average = mean(duration_mins),
            Median = median(duration_mins),
            SD = sd(duration_mins),
            range = max(duration_mins) - min(duration_mins)) %>% 
  view()
  
```

In this table, we can see the variance is much lower than before.
There are a few observations we can see here:
There are fewer causal riders than member riders; while the causal rider is willing to ride twice the amount as a member rider based on the average duration.
Also, the average and the median are not quite close to each other, which means they are not normally distributed.


#### Comparison of riding durion in minutes for casual riders and member riders by month and week.

week:
```{r}
week_duration <- total_trip_clean %>% 
  filter(duration_mins >1 & duration_mins < 1440) %>% 
  group_by(type_of_rider, day_of_week) %>%  
  summarise(number_of_rides = n(),
             average = mean(duration_mins)) %>% 
  arrange(day_of_week) %>% 
  view()
```
For casual riders, the average range is between 20mins to 27min; while the members have an average range between 12mins to 14min.
Casual riders have nearly twice the amount of duration as members.


Month:
```{r}
month_duration <- total_trip_clean %>% 
  filter(duration_mins >1 & duration_mins < 1440) %>% 
  group_by(type_of_rider, month) %>%  
  summarise(number_of_rides = n(),
             average = mean(duration_mins)) %>% 
  arrange(month) %>% 
  view()
```
The average duration by month has a similar result as the average duration by the day of week 


#### The location where they start their bikes rental.

V1 table is focusing on the exact name of the station name.
```{r}
bike_start_v1 <- total_trip_clean %>% 
  filter(duration_mins >1 & duration_mins < 1440) %>% 
  group_by(start_station_name) %>% 
  summarise(Total_number_of_start_location = n()) %>% 
  arrange(-Total_number_of_start_location) %>% 
  view()
```


V2 table is focusing on the general street/area of the bike station.
```{r}
total_trip_clean[c('split_start_loc','split_start_loc_2')] <- str_split_fixed(total_trip_clean$start_station_name, ' & ', 2)


bike_start_v2 <- total_trip_clean %>% 
  filter(duration_mins >1 & duration_mins < 1440) %>% 
  group_by(split_start_loc) %>% 
  summarise(Total_number_of_start_location = n()) 
  
bike_start_v3 <- total_trip_clean %>% 
  filter(duration_mins >1 & duration_mins < 1440) %>% 
  group_by(split_start_loc = split_start_loc_2) %>% 
  summarise(Total_number_of_start_location = n())  

bike_start_combin <- rbind(bike_start_v2,bike_start_v3)
View(bike_start_combin)

rm(bike_start_v2)
rm(bike_start_v3)
```

There are nearly 4% of riders start their rental at Clark St(The most popular); while there are only 2% of riders start their rental at Wells St(the second popular). There is more than double amount of the difference.




## Visualization
The first two graphs are comparing the number of rides on the day of week and monthly for casual riders and member riders.
#### 1.
```{r}
total_trip_clean %>%
  filter(duration_mins >1 & duration_mins < 1440) %>% 
  group_by(type_of_rider, day_of_week) %>%
  summarise(number_of_rides = n()) %>%
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = type_of_rider)) + 
  geom_bar(position = position_dodge(), stat = "identity")+
  labs(tag="1", title="Comparing number of rides on day of week", subtitle = "Aug 2021 - Jul 2022", x="Days of Week", y="Number of Rides")
```
As we can see, members tend to have a ride throughout the week. On the other hand, casual riders tend to have a ride during the weekend.

#### 2.
```{r}
total_trip_clean %>%
  filter(duration_mins >1 & duration_mins < 1440) %>% 
  group_by(type_of_rider, month) %>%
  summarise(number_of_rides = n()) %>%
  ggplot(aes(x = month, y = number_of_rides, fill = type_of_rider)) + 
  geom_bar(position = position_dodge(), stat = "identity")+
  labs(tag="2", title="Comparing number of rides of month", subtitle = "Aug 2021 - Jul 2022", x="Month", y="Number of Rides")
```
As we can see most riders tend to rent the bike between May to October due to the cold weather(below 10'C) months. (refer:https://seasonsyear.com/USA/Illinois/Chicago).

#### 3.
The third and the fourth are comparing the average duration on day of week and monthly for casual riders and member riders.
```{r}
total_trip_clean %>%
  filter(duration_mins >1 & duration_mins < 1440) %>% 
  group_by(type_of_rider, day_of_week) %>%
  summarise(average_duration = mean(duration_mins)) %>%
  ggplot(aes(x = day_of_week, y = average_duration, fill = type_of_rider)) + 
  geom_bar(position = position_dodge(), stat = "identity")+
  labs(tag="3", title="Comparing average duration on day of week", subtitle = "Aug 2021 - Jul 2022", x="Days of Week", y="Average Duration")
```

#### 4.
```{r}
total_trip_clean %>%
  filter(duration_mins >1 & duration_mins < 1440) %>% 
  group_by(type_of_rider, month) %>%
  summarise(average_duration = mean(duration_mins)) %>%
  ggplot(aes(x = month, y = average_duration, fill = type_of_rider)) + 
  geom_bar(position = position_dodge(), stat = "identity")+
  labs(tag="4", title="Comparing number of rides on month", subtitle = "Aug 2021 - Jul 2022", x="Month", y="Average Duration")
```
On both graphs, we can see those casual riders tend to ride longer than members during the whole year


Graphs 5 and 6 show the comparison of two different types of bikes, electric bikes and non-electric bikes.
#### 5.
```{r}
total_trip_clean %>%
  filter(duration_mins >1 & duration_mins < 1440) %>% 
  group_by(rideable_type, day_of_week) %>%
  summarise(number_of_rides = n()) %>%
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = rideable_type)) + 
  geom_bar(position = position_dodge(), stat = "identity")+
  labs(tag="5", title="Comparing number of rides on day of week for difference type of bikes", subtitle = "Aug 2021 - Jul 2022", x="Days of Week", y="Number of Rides")
```

#### 6.
```{r}
total_trip_clean %>%
  filter(duration_mins >1 & duration_mins < 1440) %>% 
  group_by(rideable_type, month) %>%
  summarise(number_of_rides = n()) %>%
  ggplot(aes(x = month, y = number_of_rides, fill = rideable_type)) + 
  geom_bar(position = position_dodge(), stat = "identity")+
  labs(tag="6", title="Comparing number of rides on month for different type of bike", subtitle = "Aug 2021 - Jul 2022", x="Month", y="Number of Rides")
```
Based on some of my investigation, with divvy's latest expansion on the west side of Chicago last summer, more people were riding the last summer.


The last two tables are showing which station is their starting location.
V1 table is focusing on the exact name of the station name.
V2 table is focusing on the general street/area of the bike station.
#### 7.
```{r}
bike_start_v1 <- bike_start_v1[!bike_start_v1$start_station_name=="",]

bike_start_v1%>%
  filter(Total_number_of_start_location > 37000) %>%
  mutate(start_station_name = fct_reorder(start_station_name, desc(Total_number_of_start_location))) %>%
  ggplot(aes(x = start_station_name, y = Total_number_of_start_location, fill = start_station_name)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  labs(tag="7", title="Popular bike station name in Chicago v1", subtitle = "Aug 2021 - Jul 2022", x="Station name", y="Total_amount")+
  theme(axis.text.x=element_blank())

```     
We can see that 'Streeter Dr & Grand Ave' is the most popular station to start riding compare to others.


V2 table is focusing on the general street/area of the bike station.
#### 8.
```{r}
bike_start_combin <- bike_start_combin[!bike_start_combin$split_start_loc=="",]

bike_start_combin%>%
  filter(Total_number_of_start_location > 120000) %>%
  mutate(split_start_loc = fct_reorder(split_start_loc, desc(Total_number_of_start_location))) %>% 
  ggplot(aes(x = split_start_loc, y = Total_number_of_start_location, fill = split_start_loc)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  labs(tag="8", title="Popular bike station name in Chicago v2", subtitle = "Aug 2021 - Jul 2022", x="Station name", y="Total_amount")+
  theme(axis.text.x=element_blank())
```
We can see that 'Clark st' is the most popular area to start riding compare to others.




